Modelling BCG vaccination in the UK

### 2019-12-16 rmds/05-evidence-policy-change.Rmd
completed sutherland updates

## Read formated results
## See https://github.com/seabbs/AssessBCGPolicyChange/blob/master/vignettes/paper.Rmd
## for code required to reproduce 
## Chapter is an edited version of the orginal reproducible document.
## Original results found at: https://github.com/seabbs/AssessBCGPolicyChange/blob/master/vignettes/results/paper
## Original figures found at: https://github.com/seabbs/AssessBCGPolicyChange/blob/master/docs/articles/paper_files/figure-html
##Wordcount: 4537

@@

In England and Wales, universal school-aged (at 13-14 years old) vaccination (hereafter referred to as the BCG schools scheme) was introduced after a MRC trial in the 1950s estimated BCG's effectiveness at 80% in the ethnic White UK born population.[@Hart1972] The policy remained in place until 2005, when England and Wales changed to targeted vaccination of high-risk neonates. The 2005 change in BCG vaccination policy was motivated by evidence of decreased transmission of TB, an increasing proportion of TB cases occurring in the non-UK born,[@PHE2017] and modelling evidence that suggested stopping the BCG schools scheme would have minimal long term effects on incidence rates.[@Sutherland1989] Due to the complex nature of both TB and the BCG vaccine, the ongoing impact of this change in policy is hard to directly estimate, with decision makers relying on expert opinion, evidence from surveillance data, and insight from modelling studies.

@@

\[ T = (1-d)^z < 1 \]

@@

\[ T=x+x^2+x^3+ . . = \frac{x}{1-x} \]

@@

\[ Z = \frac{xz + x^{2}2z + x^{3}3z + . .}{x+x^2+x^3+ . . }= \frac{z}{1-x} \]

@@

If we assume a constant decay rate of $d$% per year and that the next generation of secondary cases are notified $z$ years after the person who infected them, then 1 notification in year 0 will result in $(1-d)^z$ secondary notifications $z$ years later. This is therefore the relative generation size ($x$), not the *total* number of secondary notifications ($T$), as (incorrectly) stated by Sutherland et al. Therefore, to correct this I revised 1. to the following in our updated model,

\[ x = (1-d)^z < 1 \]

@@

The annual distribution of secondary notifications ($N$) was modelled by first estimating the number of secondary notifications that occurred in the current year ($i$) due to primary notifications in that year ($N^{Current}$) and then estimating how many secondary notifications occurred 5 years later ($N^{Projected}$). $N^{Current}$ was estimated using the number of primary notifications ($P$) multiplied by the number of total expected number of secondary notifications per primary notification ($T$), the proportion of secondary cases in the first generation that occurred in the first year ($f$), and the relative size of the first generation ($x$). $N^{Projected}$ was then estimated by assuming that it was equal to the number of secondary notifications, minus notifications occurring in the first year, that occurred $Z$ (the expected average interval between each primary notification and all secondary notifications) years ago. As this used the overall number of notifications from the previous time step a decay of $(1 - d)^{5-Z}$ was applied. This approach can be summarised as follows,

\[ N^{Current} = P T  f  x \]
\[ N^{Projected} = (PT - N^{Current})(1 - d)^{5-Z} \] 
\[ N_i = N^{Current}_i + N^{Projected}_{i-1} \]

@@

```{r vac-prevent-tab}
read_formated("tab-vac-prevent") %>% 
  kable(caption = "The median number (with the 2.5\\% and 97.5\\% quantiles) of vaccines required to prevent a single case of TB within 15 years in a ethnic White UK born adult vaccinated at 13 years old.",
               booktabs = TRUE) %>% 
  landscape() %>% 
  kable_styling(latex_options = "hold_position") %>% 
  kable_styling(font_size = 8) %>% 
  column_spec(2, width = "2cm")
```

@@

```{r graph-avg-add-not, fig.scap = "Annual additional (stratified into primary and secondary) notifications in 15-29 year olds from stopping the BCG schools scheme in 2006, and 2011 until 2028.", fig.cap = "Annual additional (stratified into primary and secondary) notifications in 15-29 year olds from stopping the BCG schools scheme in 2006, and 2011 until 2028. The years presented were dictated by the 5-year timestep of the model. The percentage annual decrease scenarios considered were based on those considered by the JCVI BCG subgroup. Data based scenarios and the JCVI 0% decrease scenario were not presented here as the updated transmission model could not support these scenarios. Each boxplot summarises the output of 10,000 model simulations for each scenario. Secondary notifications are reported assuming they occurred in the same year as the primary notifications that caused them.", out.extra = ""}
knitr::include_graphics("chapters/evidence-policy-change/fig-3.png")
```

```{r annual-prevent-tab}
read_formated("tab-annual_nots_prevented") %>% 
  kable(
    caption = "The median number (with the 2.5\\% and 97.5\\% quantiles) of additional annual notifications due to ending the BCG schools scheme in selected years.", booktabs = TRUE) %>% 
  landscape() %>% 
  kable_styling(latex_options = "hold_position")  %>% 
  kable_styling(font_size = 10) %>% 
  column_spec(2, width = "2cm")
```

@@

The existing method for estimating the impact of the BCG schools scheme produced uncertain estimates of the impact of ending the scheme in all years evaluated when parameter uncertainty was included. The approach used to estimate additional notifications due to transmission was found to be incorrect. Once corrected, the transmission model produced much higher estimates of additional notifications caused by ending BCG vaccination than previously reported. Further updating the model with the annual decrease in TB notifications based on both notifications and using age-specific incidence rates resulted in a decrease in the number of vaccines needed to prevent a single case in all years considered. A scenario with a 1.9% annual decrease in incidence rates was most comparable to our results based on notifications. Using this scenario, I found that the number of TB notifications arising from ending school age BCG vaccination was much greater than originally estimated using the scenario considered in Sutherland et al.

In addition to identifying that the public health impact of the change in vaccine policy was likely much larger than originally estimated, my results also provide new insights into the uncertainty of the previously published model predictions by including parameter uncertainty and measurement error and updates these predictions using newly available data. As historical data on incidence rates in the ethnic White UK born in England and Wales were not available, I considered two approaches to proxy them and investigated multiple scenarios based on those explored by the JCVI BCG subgroup. The simulation approach used here, although updated where possible, is not the most accurate method for estimating the impact of ending the BCG schools scheme as it relies on numerous assumptions based on the available knowledge in 1987 and does not account for the role of non-White and non-UK born cases. However, the strength of this work is that the estimates are based on the framework used to inform policy making. This allowed the strength of the model used in the decision-making process to be assessed once parameter uncertainty had been incorporated and for flaws in the model to be identified. This would not have been possible if the impact had been assessed using only the observed data or with an alternative model. It also allowed estimates based on updated data to be compared to historic estimates within the same framework. This would also not have been possible if a different framework had been used. As mentioned, a weakness of the model used in this study is that it did not include the whole population or age groups outside those directly affected by vaccination. Heterogeneous mixing between these groups is also likely to be an important consideration. The exclusion of these factors means that my results are likely underestimate the impact of ending the BCG schools scheme. A final limitation is that this study only considers the impact of ending the BCG schools scheme and not the impact of the introduction of the targeted neonatal vaccination program. This should be considered when evaluating the change in policy as a whole.

Little work has been done to assess the impact of the 2005 change in BCG vaccination policy or to assess the quantitative evidence used in decision making. However, multiple studies have evaluated the cost effectiveness of various BCG programs and the impact of switching between them. A cluster-randomised trial in Brazil found that BCG vaccination of those at school-age was cheaper than treatment and would prevent one TB case per 381 vaccinations even with a vaccine effectiveness of only 34% (8-53%).[@Pereira2012] This is substantially fewer than my estimate of `r vac_prevent_1_2014` (the most comparable year from our study). However, the same trial found that for regions close to the equator BCG effectiveness was low in school-age children but unchanged in neonates,[@Barreto2014a] highlighting the importance of considering the BCG vaccines reduced effectiveness near the equator when determining vaccination policy.[@Fine1995] There is also some research which supports universal re-vaccination of those at school-age, in countries with high incidence and universal vaccination of neonates, as it may be cost effective when BCG effectiveness is moderate to high.[@Barreto2014a; @Dye2013a] There is some evidence that targeted vaccination of high risk neonates maybe more cost effective than universal vaccination of neonates.[@Usher2016; @Hersh2003] However, a study in Sweden found that incidence rates in Swedish-born children increased slightly after universal vaccination of neonates was discontinued in favour of targeted vaccination.[@Romanus1992] In France, which switched from universal vaccination of neonates to targeted vaccination in 2007, it has also been shown that targeted vaccination reduced coverage in those most at risk.[@Guthmann2011] Targeted vaccination may not be more cost effective than universal vaccination when possible reductions in transmission are considered. This chapter indicated that a substantial number of cases due to transmission may be preventable if universal school-age BCG vaccination was still in place. This result is dependent on the effectiveness of BCG vaccination when given later in life, for which there is good evidence in the ethnic White UK born.[@Hart1972] I did not consider neonatal vaccination which would be less impacted by BCG's effectiveness reducing when given later in life, but may also be less likely to result in the same reductions in ongoing transmission.

@@

This chaoter has reassessed some of the evidence previously used in decision making, correcting the transmission model used, and updating the approach with new data. However, as 15 years of detailed surveillance data have been collected since the ending of the BCG schools scheme it is now possible to use regression-based approaches to estimate the direct impact on incidence rates of ending the BCG schools scheme (see Chapter \@ref(direct-eff)).[@Abbott567511] These approaches could also be used to estimate the impact of vaccinating high-risk neonates, which may outweigh any negative impacts of ending the BCG schools scheme. In addition, the development (see Chapter \@ref(model-development)), and use, of a transmission dynamic model would allow the more accurate estimation of indirect effects and the forecasting of long-term impacts (see Chapter \@ref(model-results)).

@@

* As this study used a historical approach the model used is not the accurate method for assessing the impact of ending the BCG schools scheme. However, it provides an estimate that is based on the available data and on the framework used to inform policy making. This allowed the strength of some the quantitative evidence used in the decision-making process to be assessed.

@@

* The code for the analysis contained in this Chapter can be found at: 
[doi.org/10.5281/zenodo.2583056](https://doi.org/10.5281/zenodo.2583056)^[Alternatively available from: https://github.com/seabbs/AssessBCGPolicyChange]

>>>>

## Read formated results
## See https://github.com/seabbs/AssessBCGPolicyChange/blob/master/vignettes/paper.Rmd
## for code required to reproduce 
## Chapter is an edited version of the orginal reproducible document.
## Original results and found at: https://github.com/seabbs/AssessBCGPolicyChange/blob/master/vignettes/results/paper
##Wordcount: 4537

@@

In England and Wales, universal school-aged (at 13-14 years old) vaccination (hereafter referred to as the BCG schools scheme) was introduced after a MRC trial in the 1950s estimated BCG's effectiveness at 78% in the ethnic White UK born population.[@Hart1972] The policy remained in place until 2005, when England and Wales changed to targeted vaccination of high-risk neonates. The 2005 change in BCG vaccination policy was motivated by evidence of decreased transmission of TB, an increasing proportion of TB cases occurring in the non-UK born,[@PHE2017] and modelling evidence that suggested stopping the BCG schools scheme would have minimal long term effects on incidence rates.[@Sutherland1989] Due to the complex nature of both TB and the BCG vaccine, the ongoing impact of this change in policy is hard to directly estimate, with decision makers relying on expert opinion, evidence from surveillance data, and insight from modelling studies.

@@

\begin{equation} 
  T = (1-d)^z < 1
\end{equation}

@@

\begin{equation} 
  T=x+x^2+x^3+ . . = \frac{x}{1-x}
\end{equation}

@@

\begin{equation} 
  Z = \frac{xz + x^{2}2z + x^{3}3z + . .}{x+x^2+x^3+ . . }= \frac{z}{1-x}
\end{equation}

@@

If we assume a constant decay rate of $d$% per year and that the next generation of secondary cases are notified $z$ years after the person who infected them, then 1 notification in year 0 will result in $(1-d)^z$ secondary notifications $z$ years later. This is therefore the relative generation size ($x$), not the *total* number of secondary notifications ($T$), as (incorrectly) stated by Sutherland et al. Therefore, to correct this I revised 1. to the following,

\begin{equation} 
  x = (1-d)^z < 1
\end{equation}

@@

The annual distribution of secondary notifications ($N$) was modelled by first estimating the number of secondary notifications that occurred in the current year ($i$) due to primary notifications in that year ($N^{\text{Current}}$) and then estimating how many secondary notifications occurred 5 years later ($N^{\text{Projected}}$). $N^{\text{Current}}$ was estimated using the number of primary notifications ($P$) multiplied by the number of total expected number of secondary notifications per primary notification ($T$), the proportion of secondary cases in the first generation that occurred in the first year ($f$), and the relative size of the first generation ($x$). $N^{\text{Projected}}$ was then estimated by assuming that it was equal to the number of secondary notifications, minus notifications occurring in the first year, that occurred $Z$ (the expected average interval between each primary notification and all secondary notifications) years ago. As this used the overall number of notifications from the previous time step a decay of $(1 - d)^{5-Z}$ was applied. This approach can be summarised as follows,

\begin{equation} 
  N^{\text{Current}} = P T  f  x 
\end{equation}

\begin{equation} 
  N^{\text{Projected}} = (PT - N^{\text{Current}})(1 - d)^{5-Z}
\end{equation}

\begin{equation} 
  N_i = N^{\text{Current}}_i + N^{\text{Projected}}_{i-1}
\end{equation}

@@

```{r vac-prevent-tab}
read_formated("tab-vac-prevent") %>% 
  kable(caption = "The median number (with the 2.5\\% and 97.5\\% quantiles) of vaccines required to prevent a single case of TB within 15 years in a ethnic White UK born adult vaccinated at 13 years old. The percentage annual decrease scenarios considered were based on those considered by the JCVI BCG subgroup, with the addition of a scenario using aggregate notification data and a scenario using estimates of age-specific incidence rates in the UK born.",
               booktabs = TRUE) %>% 
  landscape() %>% 
  kable_styling(latex_options = "hold_position") %>% 
  kable_styling(font_size = 8) %>% 
  column_spec(2, width = "2cm")
```

@@

```{r graph-avg-add-not, fig.scap = "Annual additional (stratified into primary and secondary) notifications in 15-29 year olds from stopping the BCG schools scheme in 2006, and 2011 until 2028.", fig.cap = "Annual additional (stratified into primary and secondary) notifications in 15-29 year olds from stopping the BCG schools scheme in 2006, and 2011 until 2028. The years presented were dictated by the 5-year timestep of the model. The percentage annual decrease scenarios considered were based on those considered by the JCVI BCG subgroup. Data based scenarios and the JCVI 0\\% decrease scenario were not presented here as the updated transmission model could not support these scenarios. Each boxplot summarises the output of 10,000 model simulations for each scenario. Secondary notifications are reported assuming they occurred in the same year as the primary notifications that caused them.", out.extra = ""}
knitr::include_graphics("chapters/evidence-policy-change/fig-3.png")
```

```{r annual-prevent-tab}
read_formated("tab-annual_nots_prevented") %>% 
  kable(
    caption = "The median number (with the 2.5\\% and 97.5\\% quantiles) of additional annual notifications due to ending the BCG schools scheme in selected years. The percentage annual decrease scenarios considered were based on those considered by the JCVI BCG subgroup. Data based scenarios and the JCVI 0\\% decrease scenario were not presented here as the updated transmission model could not support these scenarios.", booktabs = TRUE) %>% 
  landscape() %>% 
  kable_styling(latex_options = "hold_position")  %>% 
  kable_styling(font_size = 10) %>% 
  column_spec(2, width = "2cm")
```

@@

The existing method for estimating the impact of the BCG schools scheme produced uncertain estimates of the impact of ending the scheme in all years evaluated when parameter uncertainty was included. The approach used to estimate additional notifications due to transmission was found to be incorrect. Once corrected, the transmission model produced much higher estimates of additional notifications caused by ending BCG vaccination than previously reported. Further updating the model with the annual decrease in TB notifications based on both notifications and using age-specific incidence rates resulted in a decrease in the number of vaccines needed to prevent a single case in all years considered. A scenario with a 1.9% annual decrease in incidence rates was most comparable to the results based on notifications. Using this scenario, I found that the number of TB notifications arising from ending school age BCG vaccination was much greater than originally estimated using the scenario considered in Sutherland et al.

In addition to identifying that the public health impact of the change in vaccine policy was likely much larger than originally estimated, my results also provide new insights into the uncertainty of the previously published model predictions by including parameter uncertainty and measurement error and updates these predictions using newly available data. As historical data on incidence rates in the ethnic White UK born in England and Wales were not available, I considered two approaches to proxy them and investigated multiple scenarios based on those explored by the JCVI BCG subgroup. The simulation approach used here, although updated where possible, is not the most accurate method for estimating the impact of ending the BCG schools scheme as it relies on numerous assumptions based on the available knowledge in 1987 and does not account for the role of non-White and non-UK born cases. However, the strength of this work is that the estimates are based on the framework used to inform policy making. This allowed the strength of the model used in the decision-making process to be assessed once parameter uncertainty had been incorporated and for flaws in the model to be identified. This would not have been possible if the impact had been assessed using only the observed data or with an alternative model. It also allowed estimates based on updated data to be compared to historic estimates within the same framework. This would also not have been possible if a different framework had been used. As mentioned, a weakness of the model used in this study is that it did not include the whole population or age groups outside those directly affected by vaccination. Heterogeneous mixing between these groups is also likely to be an important consideration. The exclusion of these factors means that my results are likely to underestimate the impact of ending the BCG schools scheme. A final limitation is that this study only considers the impact of ending the BCG schools scheme and not the impact of the introduction of the targeted neonatal vaccination program. This should be considered when evaluating the change in policy as a whole.

Little work has been done to assess the impact of the 2005 change in BCG vaccination policy or to assess the quantitative evidence used in decision making. However, multiple studies have evaluated the cost effectiveness of various BCG programs and the impact of switching between them. A cluster-randomised trial in Brazil found that BCG vaccination of those at school-age was cheaper than treatment and would prevent one TB case per 381 vaccinations even with a vaccine effectiveness of only 34% (8-53%).[@Pereira2012] This is substantially fewer than my estimate of `r vac_prevent_1_2014` (the most comparable year from my results). However, the same trial found that for regions close to the equator BCG effectiveness was low in school-age children but unchanged in neonates,[@Barreto2014a] highlighting the importance of considering the BCG vaccines reduced effectiveness near the equator when determining vaccination policy.[@Fine1995] There is also some research which supports universal re-vaccination of those at school-age, in countries with high incidence and universal vaccination of neonates, as it may be cost effective when BCG effectiveness is moderate to high.[@Barreto2014a; @Dye2013a] There is some evidence that targeted vaccination of high risk neonates maybe more cost effective than universal vaccination of neonates.[@Usher2016; @Hersh2003] However, a study in Sweden found that incidence rates in Swedish-born children increased slightly after universal vaccination of neonates was discontinued in favour of targeted vaccination.[@Romanus1992] In France, which switched from universal vaccination of neonates to targeted vaccination in 2007, it has also been shown that targeted vaccination reduced coverage in those most at risk.[@Guthmann2011] Targeted vaccination may not be more cost effective than universal vaccination when possible reductions in transmission are considered. This chapter indicated that a substantial number of cases due to transmission may be preventable if universal school-age BCG vaccination was still in place. This result is dependent on the effectiveness of BCG vaccination when given later in life, for which there is good evidence in the ethnic White UK born.[@Hart1972] I did not consider neonatal vaccination which would be less impacted by BCG's effectiveness reducing when given later in life, but may also be less likely to result in the same reductions in ongoing transmission.

@@

This chapter has reassessed some of the evidence previously used in decision making, correcting the transmission model used, and updating the approach with new data. However, as 15 years of detailed surveillance data have been collected since the ending of the BCG schools scheme it is now possible to use regression-based approaches to estimate the direct impact on incidence rates of ending the BCG schools scheme (see Chapter \@ref(direct-eff)).[@Abbott567511] These approaches could also be used to estimate the impact of vaccinating high-risk neonates, which may outweigh any negative impacts of ending the BCG schools scheme. In addition, the development (see Chapter \@ref(model-development)), and use, of a transmission dynamic model would allow the more accurate estimation of indirect effects and the forecasting of long-term impacts (see Chapter \@ref(model-results)).

@@

* As this study used a historical approach the model used is not the most accurate method for assessing the impact of ending the BCG schools scheme. However, it provides an estimate that is based on the available data and on the framework used to inform policy making. This allowed the strength of some the quantitative evidence used in the decision-making process to be assessed.

@@

* The code for the analysis contained in this chapter can be found at: 
[doi.org/10.5281/zenodo.2583056](https://doi.org/10.5281/zenodo.2583056)^[Alternatively available from: https://github.com/seabbs/AssessBCGPolicyChange]
